Columnar VCF for Biobank scale datasets using Zarr

### 2024-04-10 paper.tex
Pass through whole-matrix compute section

\subsection{Calculating with the genotype matrix}
Storing genetic variation data compactly is important, but it is also
important that we can analyse the data efficiently. Bioinformatics 
workflows tend to emphasise text files and command line utilities 
that consume and produce text~\citep[e.g.][]{buffalo2015bioinformatics}. 
Thus, many tools that compress VCF data provide a command line 
utility with a query language to restrict the records
examined, perform some pre-specified calculations and finally 
output some text, typically VCF or tab/comma separated values
[TODO check these citations]~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
These pre-defined calculations are by necessity limited in scope, however,
and the volumes of text involved in Biobank scale datasets
make the classical approach of custom
analyses via Unix utilities in pipelines prohibitively slow. Thus, 
methods have begun to provide Application Programming Interfaces
(APIs), providing efficient access to genotype and other VCF data
[TODO check these citations]~\cite[e.g.][]{lefaive2021sparse,
wertenbroek2022xsi,zhang2023gbc}.

To demonstrate the accessibility of genotype data and efficiency with 
which calculations can be performed under the different formats,
we use the \texttt{bcftools +af-dist} plugin as an example of such a whole-matrix
operation. 
% The details of the \texttt{af-dist} operation are not important:
% as an example of a whole-matrix operation. 
We chose this particular operation for several
reasons. First, it is a simple but non-trival calculation that requires 
examining every element in the genotype matrix.
Secondly, it produces a small volume of output (a table of
deviations from Hardy-Weinberg expectations in ten allele frequency
bins) and therefore the time spent outputting results is negligible.
Finally, it has an efficient implementation written using the 
\texttt{htslib} C API~\citep{bonfield2021htslib}, distributed as 
\texttt{bcftools} plugin.

@@

Figure~\ref{fig-whole-matrix-compute} shows the results of 
running \texttt{bcftools +af-dist} and equivalent operations 
on the data of Figure~\ref{fig-data-storage}. We can see 
that for $10^6$ samples, running \texttt{bcftools +af-dist}
requires 20 hours (note that this is run on a BCF file, for 
efficiency), which can be regarded as a reasonable baseline.
Because genozip does not provide an API, we used the classical
approach of piping the VCF text that it outputs into 
\texttt{bcftools +af-dist}. The volumes of data involved here
make this an expensive approach.
[TODO discuss Zarr-python and Savvy C++ implementations]

% We illustrate the discussion by the afdist operation from
% the bcftools plugins package. This is perhaps a contrived example, but
% we chose it because it requires examining the entire genotype matrix
% and there is an efficient implementation available in bcftools.
% It is also a good example because the standard approach of
% divvying up the genome into chunks via HPC schedulers or workflow
% managers is quite tricky.
% We discuss the software development aspect, and how we developed
% the afdist-savvy program. We discuss the inherent limitations

>>>>

\subsection{Calculating with the genotype matrix}
Storing genetic variation data compactly is important, but it is also
important that we can analyse the data efficiently. Bioinformatics 
workflows tend to emphasise text files and command line utilities 
that consume and produce text~\citep[e.g.][]{buffalo2015bioinformatics}. 
Thus, many tools that compress VCF data provide a command line 
utility with a query language to restrict the records
examined, perform some pre-specified calculations and finally 
output some text, typically VCF or tab/comma separated values
[TODO check these citations]~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
These pre-defined calculations are by necessity limited in scope, however,
and the volumes of text involved in Biobank scale datasets
make the classical approach of custom
analyses via Unix utilities in pipelines prohibitively slow. Thus, 
methods have begun to provide Application Programming Interfaces
(APIs), providing efficient access to genotype and other VCF data
[TODO check these citations]~\cite[e.g.][]{kelleher2013processing,lefaive2021sparse,
wertenbroek2022xsi,zhang2023gbc}. By providing programmatic access,
the data can be retrieved from storage, decoded and then analysed
in the same memory space, avoiding the need for copying
the decoded data and inter-process communication. [TODO some refs
on zero-copy semantics and things like Arrow?]

To demonstrate the accessibility of genotype data and efficiency with 
which calculations can be performed under the different formats,
we use the \texttt{bcftools +af-dist} plugin as an example of such a whole-matrix
computation. 
% The details of the \texttt{af-dist} operation are not important:
% as an example of a whole-matrix operation. 
We chose this particular operation for several
reasons. First, it is a simple but non-trival calculation that requires 
examining every element in the genotype matrix.
Secondly, it produces a small volume of output (a table of
deviations from Hardy-Weinberg expectations in ten allele frequency
bins) and therefore the time spent outputting results is negligible.
Finally, it has an efficient implementation written using the 
\texttt{htslib} C API~\citep{bonfield2021htslib}, distributed as 
a \texttt{bcftools} plugin.

@@

Figure~\ref{fig-whole-matrix-compute} shows the results of 
running \texttt{bcftools +af-dist} and equivalent operations 
on the data of Figure~\ref{fig-data-storage}. There is a large
difference in the time required (note the log-log scale). 
The slowest approach uses Genozip. Because genozip does not
provide an API and only outputs VCF text, the best approach available 
is to pipe its output into \texttt{bcftools +af-dist}. Thus,
not only must we decode the data from the genozip format, we must 
then parse the large volumes of VCF text and then finally do the 
actual calculation. Running \texttt{bcftools +af-dist} directly
on the gzipped VCF text incurs essentially the same cost, but
the decompression cost is lower and without the interprocess 
communication overhead. Using a BCF file is somewhat faster,
because the packed binary format can be read into the internal 
data structures used by the \texttt{htslib} API more efficiently.

The data shown in Figure~\ref{fig-whole-matrix-compute} for Zarr and Savvy
is based on custom programs written using their respective APIs
to implement the \texttt{af-dist} operation. The Zarr program uses
the Zarr-Python package to iterate over the decoded chunks of the 
genotype matrix and classifies genotypes within a chunk using a 14 line Python
function, accelerated using the Numba JIT compliler~\cite{lam2015numba}.
The allele frequencies and genotype counts are then analysed to produce 
the final counts within the allele frequency bins with 9 lines of 
Python using NumPy~\cite{harris2020array} functions. Remarkably, this 
short and simple Python program is substantially faster than the 
equivalent compiled C using \texttt{htslib} APIs on BCF (6 hours
vs 20 hours for 1 million samples). The fastest method is the 
C++ program written using the Savvy API. This would largely seem
to be due to Savvy's excellent genotype decoding performance
[TODO add fig for decode perf].
